---
title: "gov51_ps3"
format: pdf
editor: visual
---

## Joshua Stewart

GitHub Link: \_\_\_\_\_

## 1 Setup and Data Exploration (10 Points)

### 1.1 Load packages and data

Load the tidyverse, glmnet, and knitr packages. Then load the COMPAS dataset from data/compas_clean.csv.

```{r}
library(tidyverse)
library(glmnet)
library(knitr)

compas <- read_csv("data/raw/compas_clean.csv")
```

### **1.2 How many observations and variables are in the dataset? What is the unit of
observation?**

```{r}
nrow(compas)
ncol(compas)
head(compas)
```

There are 7,214 observations and 17 variables in the data set. Each observation is a person.

### **1.3 Examine the outcome variable**

**The variable recidivism equals 1 if the person was rearrested within two years and 0 otherwise.
(a) What is the overall recidivism rate in the data?**

```{r}
mean(compas$recidivism, na.rm = TRUE)
```

Approximately 45.1% of individuals recidivated.

**(b) Is this outcome continuous or binary? Why does this matter for how we interpret predictions from a linear model?**

The recidivism variable is binary: 0 if the individual has not recidivized and 1 if they have. This matters for how we interpret predictions from a linear model because in linear models, when the outcome is binary, the prediction is interpreted as a probability. That said, linear models can produce values above 1 and below 0, which would not make sense for this scenario.

### 1.4 Summary statistics

**Create a summary statistics table showing the mean and standard deviation of all variables in
the dataset. Briefly note which variables have the most variation and which have the least.**

```{r}
library(dplyr)

compas |> 
  summarise(across(everything(),
                   list(mean = ~mean(.x, na.rm = TRUE),
                        sd = ~sd(.x, na.rm = TRUE))))
```

The priors_count variable has the most variation (SD = 4.88), with age following behind (SD = 11.89). That said, the variation in age is expected. \

The variables with the least variation are the race variables. For example, the native american and asian identities have a standard deviation of 0.05 and 0.066. That said, I assume ti is because they are binary indicators that appear less frequently in the sample.

### 1.5 Explore key predictors

**Create a bar chart showing the recidivism rate by number of prior offenses (priors_count).
Group priors_count into categories: 0, 1–2, 3–5, 6–10, and 11+. What pattern do you see?**

```{r}
compas_gr <- compas |>
  mutate(priors_group = case_when(priors_count == 0 ~ "0",
                                  priors_count %in% 1:2 ~ "1–2",
                                  priors_count %in% 3:5 ~ "3–5",
                                  priors_count %in% 6:10 ~ "6–10",
                                  priors_count >= 11 ~ "11+"))
summary_table <- compas_gr |>
  group_by(priors_group) |>
  summarize(recid_rate = mean(recidivism, na.rm = TRUE))

ggplot(summary_table, aes(x = priors_group, y = recid_rate)) +
  geom_col(fill = "pink") +
  labs(x = "Number of Prior Offenses",
       y = "Recidivism Rate",
       title = "Recidivism Rate by Number of Prior Offense")
```

## 2 OLS Baseline (15 points)

### 2.1 Train/test split

**Set your seed to 51 for reproducibility. Split the data into a training set (80%) and a test set
(20%).**

```{r}
set.seed(51)
n <- nrow(compas)
train_idx <- sample(1:n, size = floor(0.8 * n))
train <- compas[train_idx, ]
test <- compas[-train_idx, ]

nrow(train)
nrow(test)
```

**How many observations are in each set?**

There are 5,771 observations in the the train data set and 1,443 in the test data set.

### 2.2 Fit OLS with main effects only

**Fit an OLS regression of recidivism on all other variables in the training data using lm(recidivism \~ ., data = train). This gives you a model with 16 predictors.**

```{r}
summary(lm(recidivism ~ ., data = train))
```

**(a) Which predictors are statistically significant at the 5% level?**

Predictors are statistically significant at the 5% level if they are less than 0.05. In this data set, the following predictors fulfill this requirement: age (p \< 2e-16), female (p = 0.00), priors_count (p \< 2e-16), felony (0.00), has juvenile misdemeanor (p = 0.04), and age between 25-45 (p = 0.00).

**(b) Interpret the coefficient on priors_count in one sentence.**

The coefficient on priors_count (0.03) means that each additional prior offense is associated with approximately a 3 percentage point increase in probability of recidivism, holding all other variables constant.

### 2.3 In-sample vs. out-of-sample RMSE

**Root Mean Squared Error (RMSE) measures average prediction error.**

**(a) Compute the in-sample RMSE (predictions on the training data).**

```{r}
model <- lm(recidivism ~ ., data = train)
train_pred <- predict(model, newdata = train)

train_rmse <- sqrt(mean((train_pred - train$recidivism)^2))
train_rmse
```

**(b) Compute the out-of-sample RMSE (predictions on the test data).**

```{r}
test_pred <- predict(model, newdata = test)

test_rmse <- sqrt(mean((test_pred - test$recidivism)^2))
test_rmse
```

**(c) How large is the gap between in-sample and out-of-sample RMSE? Is OLS overfitting
with 16 predictors?**

```{r}
test_rmse - train_rmse
```

The difference between in-sample (RMSE = 0.46) and
