---
title: "gov51_ps3"
format: pdf
editor: visual
---

## Joshua Stewart

GitHub Link: <https://github.com/JoshuaMcKoy/gov51_ps3>

## 1 Setup and Data Exploration (10 Points)

### 1.1 Load packages and data

Load the tidyverse, glmnet, and knitr packages. Then load the COMPAS dataset from data/compas_clean.csv.

```{r}
library(tidyverse)
library(glmnet)
library(knitr)

compas <- read_csv("../data/raw/compas_clean.csv")
```

### 1.2 How many observations and variables are in the dataset? What is the unit of observation?

```{r}
nrow(compas)
ncol(compas)
head(compas)
```

There are 7,214 observations and 17 variables in the data set. Each observation is a person.

### **1.3 Examine the outcome variable**

**The variable recidivism equals 1 if the person was rearrested within two years and 0 otherwise. (a) What is the overall recidivism rate in the data?**

```{r}
mean(compas$recidivism, na.rm = TRUE)
```

Approximately 45.1% of individuals recidivated.

**(b) Is this outcome continuous or binary? Why does this matter for how we interpret predictions from a linear model?**

The recidivism variable is binary: 0 if the individual has not recidivized and 1 if they have. This matters for how we interpret predictions from a linear model because in linear models, when the outcome is binary, the prediction is interpreted as a probability. That said, linear models can produce values above 1 and below 0, which would not make sense for this scenario.

### 1.4 Summary statistics

**Create a summary statistics table showing the mean and standard deviation of all variables in the dataset. Briefly note which variables have the most variation and which have the least.**

```{r}
library(dplyr)

compas |> 
  summarise(across(everything(),
                   list(mean = ~mean(.x, na.rm = TRUE),
                        sd = ~sd(.x, na.rm = TRUE))))
```

The priors_count variable has the most variation (SD = 4.88), with age following behind (SD = 11.89). That said, the variation in age is expected.\

The variables with the least variation are the race variables. For example, the native american and asian identities have a standard deviation of 0.05 and 0.066. That said, I assume ti is because they are binary indicators that appear less frequently in the sample.

### 1.5 Explore key predictors

**Create a bar chart showing the recidivism rate by number of prior offenses (priors_count). Group priors_count into categories: 0, 1–2, 3–5, 6–10, and 11+. What pattern do you see?**

```{r}
compas_gr <- compas |>
  mutate(priors_group = case_when(priors_count == 0 ~ "0",
                                  priors_count %in% 1:2 ~ "1–2",
                                  priors_count %in% 3:5 ~ "3–5",
                                  priors_count %in% 6:10 ~ "6–10",
                                  priors_count >= 11 ~ "11+"))
summary_table <- compas_gr |>
  group_by(priors_group) |>
  summarize(recid_rate = mean(recidivism, na.rm = TRUE))

ggplot(summary_table, aes(x = priors_group, y = recid_rate)) +
  geom_col(fill = "pink") +
  labs(x = "Number of Prior Offenses",
       y = "Recidivism Rate",
       title = "Recidivism Rate by Number of Prior Offense")
```

## 2 OLS Baseline (15 points)

### 2.1 Train/test split

**Set your seed to 51 for reproducibility. Split the data into a training set (80%) and a test set (20%).**

```{r}
set.seed(51)
n <- nrow(compas)
train_idx <- sample(1:n, size = floor(0.8 * n))
train <- compas[train_idx, ]
test <- compas[-train_idx, ]

nrow(train)
nrow(test)
```

**How many observations are in each set?**

There are 5,771 observations in the the train data set and 1,443 in the test data set.

### 2.2 Fit OLS with main effects only

**Fit an OLS regression of recidivism on all other variables in the training data using lm(recidivism \~ ., data = train). This gives you a model with 16 predictors.**

```{r}
summary(lm(recidivism ~ ., data = train))
```

**(a) Which predictors are statistically significant at the 5% level?**

Predictors are statistically significant at the 5% level if they are less than 0.05. In this data set, the following predictors fulfill this requirement: age (p \< 2e-16), female (p = 0.00), priors_count (p \< 2e-16), felony (0.00), has juvenile misdemeanor (p = 0.04), and age between 25-45 (p = 0.00).

**(b) Interpret the coefficient on priors_count in one sentence.**

The coefficient on priors_count (0.03) means that each additional prior offense is associated with approximately a 3 percentage point increase in probability of recidivism, holding all other variables constant.

### 2.3 In-sample vs. out-of-sample RMSE

**Root Mean Squared Error (RMSE) measures average prediction error.**

**(a) Compute the in-sample RMSE (predictions on the training data).**

```{r}
model <- lm(recidivism ~ ., data = train)
train_pred <- predict(model, newdata = train)

train_rmse <- sqrt(mean((train_pred - train$recidivism)^2))
train_rmse
```

**(b) Compute the out-of-sample RMSE (predictions on the test data).**

```{r}
test_pred <- predict(model, newdata = test)

test_rmse <- sqrt(mean((test_pred - test$recidivism)^2))
test_rmse
```

**(c) How large is the gap between in-sample and out-of-sample RMSE? Is OLS overfitting with 16 predictors?**

```{r}
test_rmse - train_rmse
```

The difference between in-sample RMSE (0.46) and the out-of-sample RMSE (0.46) is 0.0016. Given that the model generalizes well to the out-of-sample estimartes, OLS is not overfitting the data.

### 2.4 Cross-validation by hand

**A single train/test split depends on which observations happened to land in each group. Cross-validation averages over multiple splits to give a more stable estimate of out-of-sample performance.**

**Implement 5-fold cross-validation for the simple OLS model:
1. Randomly assign each training observation to one of 5 folds.
2. For each fold 𝑘 = 1, … , 5: fit OLS on the other 4 folds, predict fold 𝑘, compute RMSE.
3. Average the 5 RMSE values**

```{r}
set.seed(51)
folds <- sample(rep(1:5, length.out = nrow(train)))

rmse_cv <- numeric(5)
for (k in 1:5) {train_k <- train[folds != k, ]
  test_k <- train[folds == k, ]
  model_k <- lm(recidivism ~ ., data = train_k)
  pred_k <- predict(model_k, newdata = test_k)
  rmse_cv[k] <- sqrt(mean((test_k$recidivism - pred_k)^2))}

mean(rmse_cv)
```

**(a) What is the 5-fold CV estimate of RMSE? How does it compare to your single-split
out-of-sample RMSE?**

The 5-fold CV estimate of RMSE is 0.46. This matches our single-split out-of-sample RMSE.

**(b) In your own words, why is cross-validation more reliable than a single train/test split?**

Cross-validation is more reliable than a single train/test split because it averages prediction error across multiple different splits of the data, instead of relying on a single train/test split. This reduces the impact of random variation in how the data is divided.

**(c) When we use cv.glmnet() later, it does this same process internally to choose the
penalty parameter 𝜆. Why is cross-validation the right tool for choosing 𝜆?**

Cross-validation is the right tool for choosing λ because it allows us to choose the value of λ that minimizes prediction error on out-of-sample data, helping balance overfitting and underfitting.

## 3 The Kitchen Sink: What Happens When You Add Too Many Predictors? (15 points)

**The simple OLS model uses 16 main effects. But maybe interactions between variables would
help. What if being young AND having prior offenses is different from the sum of each effect
alone? Let’s find out what happens when we throw in everything.**

### 3.1 Build the expanded model

The formula below creates all pairwise interactions plus squared terms for the continuous
variables. This expands our feature set from 16 to about 141 predictors.

```{r}
f_full <- recidivism ~ (.)^2 +
I(age^2) + I(priors_count^2) +
I(juv_fel_count^2) + I(juv_misd_count^2) +
I(juv_other_count^2)

X_train <- model.matrix(f_full, data = train)[, -1]
X_test <- model.matrix(f_full, data = test)[, -1]
y_train <- train$recidivism
y_test <- test$recidivism

ncol(X_train)
```

**(a) How many predictors does this expanded model have?**

This expanded model has 141 predictors. **(b) What is the ratio of training observations to predictors? (Compare to the simple model.)**

```{r}
5771/141
5771/16
```

The ratio of training observations in the expanded model is 40.93:1, while it is 360.69:1 in the simple model.

### 3.2 Fit OLS with the kitchen sink

**Fit OLS on the expanded feature set: ols_full \<- lm(f_full, data = train).**

```{r}
ols_full <- lm(f_full, data = train)
```

**(a) Compute the in-sample RMSE for this model.**

```{r}
pred_train <- predict(ols_full, newdata = train)

rmse_train_full <- sqrt(mean((train$recidivism - pred_train)^2))
rmse_train_full
```

**(b) Compute the out-of-sample RMSE.**

```{r}
pred_test <- predict(ols_full, newdata = test)

rmse_test_full <- sqrt(mean((test$recidivism - pred_test)^2))
rmse_test_full
```

**(c) How does the in-sample/out-of-sample gap compare to the simple 19-predictor OLS? What happened and why?**

```{r}
rmse_test_full-rmse_train_full
```

The in-sample/out-of-sample gap is larger in the expanded model (0.04) than in the simple model (0.0016). This demonstrates that the newer model might be overfitting the data the training data, increasing training data fit but decreasing predicitve performance with the test sample.

### 3.3 The overfitting lesson

**In 2–3 sentences, explain why adding 180 interaction terms made the model’s in-sample fit better but its out-of-sample predictions worse. What is this phenomenon called?**

Adding interaction terms increases the model's complexity, which imrpoves the in-sample fit by capturing more patterns of the training data. That said, this leads the model to fit noise rather than signals, resulting in worse performance iwth the out-of-sample test data. This is called overfitting the training data.

## 4 Ridge Regression: Shrink Everything (15 points)

**Ridge regression adds an L2 penalty that shrinks all coefficients toward zero. Larger 𝜆 means
more shrinkage. The key idea: Ridge trades a small amount of bias for a large reduction in
variance.**

### 4.1 Fit Ridge with cross-validation

**Use cv.glmnet() with alpha = 0 to fit Ridge regression on the expanded training data (the
X_train matrix with \~141 features). Use 10-fold cross-validation (the default) to select the
optimal 𝜆.**

```{r}
set.seed(51)
cv_ridge <- cv.glmnet(X_train, y_train, alpha = 0, nfolds = 10)
```

**(a) What is the optimal 𝜆 selected by cross-validation? (Use lambda.min.)**

```{r}
cv_ridge$lambda.min
```

The optimal **𝜆** selected by cross-validation is 0.014.

**(b) Plot the cross-validation curve using plot(cv_ridge). What does this plot show?**

```{r}
plot(cv_ridge)
```

The plot shows the cross-validated mean squared error as a function of λ. The error decreases as λ increases and then levels off, indicating that reducing regularization improves model fit up to a point. The optimal λ (lambda.min = 0.014) minimizes the cross-validated error.

### 4.2 Ridge RMSE

**Compute the out-of-sample RMSE for Ridge. Compare to both the simple OLS (16 vars) and the kitchen-sink OLS (\~141 vars). What happened?**

```{r}
pred_ridge <- predict(cv_ridge, newx = X_test, s = "lambda.min")

rmse_ridge <- sqrt(mean((y_test - pred_ridge)^2))
rmse_ridge
```

The out-of-sample RMSE for Ridge (0.47) is a little less accurate than simple OLS (0.46) and a bit more accurate than the kitchen-sink OLS (0.49). Ridge improves the performance by shrinking coefficients and reducing variance, allowing it to handle the large number of predictors more effectively, although it does not outperform the simpler OLS model.

### 4.3 Did Ridge set any coefficients to zero?

**Extract the Ridge coefficients using coef(cv_ridge, s = "lambda.min"). Did Ridge eliminate any variables? Why or why not?**

```{r}
coef(cv_ridge, s = "lambda.min")
```

Ridge did not eliminate variables, though some of them naturally have reduced to zero due to factors unrelated to the ridge penalty.

## **5 LASSO Regression: Shrink and Select (15 points)**

**LASSO (Least Absolute Shrinkage and Selection Operator) uses an L1 penalty. Unlike Ridge, LASSO can shrink coeﬀicients all the way to exactly zero, effectively performing variable selection.**		

### 5.1 Fit LASSO with cross-validation

**Use cv.glmnet() with alpha = 1 on the expanded training data. Remember to set your seed
first for reproducibility.**

```{r}
set.seed(51)
cv_lasso <- cv.glmnet(X_train, y_train, alpha = 1, nfolds = 10)
```

**(a) What is the optimal 𝜆?**

```{r}
cv_lasso$lambda.min
```

The optimal **𝜆** 0.0077.

**(b) Plot the cross-validation curve.**

```{r}
plot(cv_lasso)
```

### 5.2 LASSO variable selection

**Extract the LASSO coefficients at lambda.min.
(a) How many of the \~141 variables survived (non-zero coefficient)? How many were eliminated?**

```{r}
lasso_coef <- coef(cv_lasso, s = "lambda.min")

coef_values <- as.vector(lasso_coef)
coef_names <- rownames(lasso_coef)

nonzero <- coef_names[coef_values != 0]

nonzero

sum(coef_values != 0)
sum(coef_values == 0)
```

Out of approximately 141 predictors, 28 have non-zero coefficients and 114 are eliminated/set to zero.

**(b) List a few of the surviving interaction terms. Do they make substantive sense?**

Some surviving interaction terms include age:female, priors_count:age_25_45, and juv_fel_count:age_under_25. I would say that these make substantive sense because they allow the effect of variables such as prior offenses or juvenile history on recidivism to vary across demographic groups like age or gender.

**(c) Baker (2025) argues that LASSO reduces “subjective discretion” in model building. How does what you see here illustrate that argument?**

This illustrates Baker’s argument because LASSO automatically selects a subset of variables based on predictive performance, rather than individually choosing which variables or interactions to include. From my understanding, by retaining only interactions that contribute most to prediction, LASSO reduces subjective discretion.

### **5.3 LASSO RMSE**

**Compute the out-of-sample RMSE for LASSO. How does it compare to OLS (both versions) and Ridge?**

```{r}
pred_lasso <- predict(cv_lasso, newx = X_test, s = "lambda.min")
rmse_lasso <- sqrt(mean((y_test - pred_lasso)^2))
rmse_lasso
```

The LASSO RMSE is 0.46, which is incredibly similar to the simple OLS RMSE (0.46) and smaller than the kitchen-sink OLS RMSE (0.40) and the Ridge RMSE (0.47). Thus, LASSO improves the prediction by eliminating the less necessary variables.

## 6 Model Comparison and Interpretation (30 points)

### **6.1 Elastic Net**

**Fit an Elastic Net with alpha = 0.5 on the expanded training data. Remember to set your seed first. Compute its out-of-sample RMSE and count how many variables it retains.**

```{r}
set.seed(51)
cv_enet <- cv.glmnet(X_train, y_train, alpha = 0.5, nfolds = 10)

pred_enet <- predict(cv_enet, newx = X_test, s = "lambda.min")

rmse_enet <- sqrt(mean((y_test - pred_enet)^2))
rmse_enet

enet_coef <- coef(cv_enet, s = "lambda.min")

coef_values <- as.vector(enet_coef)

sum(coef_values[-1] != 0)
```

The out-of-sample RMSE is approximately 0.46, and the model retains 43 variables.

### 6.2 The big comparison table

**Fill in this table with your results:**

```{r}
results <- data.frame(Model = c("OLS (main effects)", 
                                "OLS (kitchen sink)", 
                                "Ridge", 
                                "LASSO", 
                                "Elastic Net"),
                      Predictors = c(16, 
                                     141, 
                                     141, 
                                     28, 
                                     sum(as.vector(coef(cv_enet, s = "lambda.min"))[-1] != 0)),
                      "Out-of-Sample RMSE" = c(test_rmse, 
                                               rmse_test_full, 
                                               rmse_ridge, 
                                               rmse_lasso, 
                                               rmse_enet))
results
```

**(a) Which model has the lowest out-of-sample RMSE?**

The elastic net model has the lowest out-of-sample RMSE (0.458).

**(b) Which model has the best in-sample fit? Is that the same model?**

The kitchen-sink OLS model has the best in-sample fit because it has the lowest training error due to its large number of predictors. This is not the same model with the lowest out-of-sample RMSE, as it overfits the training data and performs worse on new data.

**(c) In 2–3 sentences, explain what this table teaches about the bias-variance tradeof.**

This table teaches us that models with more predictors may decrease variance at the expense of increasing bias because the system overfits to the idiosyncracies of the training data, making it less useful in actual prediction. On the other hand, allowing for a little bit of bias can drastically increase the predictive power of the model.

### 6.3 Baker article connection

**In Baker (2025), two experts in the Halliburton securities case selected different peer firms and reached different conclusions about whether a stock drop was statistically significant.
(a) How did LASSO help narrow the disagreement between dueling experts?**

By automatically shrinking some coefficients to zero and keeping only the most predictive variables, LASSO reduces the ability of different experts to reach different conclusions based on different model specifications, instead of the analysts' subjective choices.

**(b) Based on your experience with LASSO in this problem set, what aspect of the traditional OLS approach gives the analyst the most room for manipulation?**

Based on my experience with LASSO, the traditional OLS approach gives the analysts the most room for manipulation in variable/interaction selection. As seen in this problem set, adding or removing predictors can substantially change the results, allowing analysts to influence conclusions through subjective model specification.

### 6.4 Prediction vs. explanation

**We used these models to predict recidivism. Suppose instead you wanted to know whether
priors_count causes higher recidivism.
(a) Would the LASSO coefficient on priors_count answer this causal question? Why or why not?**

The LASSO coefficient on priors_count would not answer this causal question. First, LASSO is designed for prediction, not causality. Second, the estimated relationship may reflect correlations driven by confounding variables rather than a true causal effect.

**(b) What is the difference between using regression for prediction versus using it for causal inference?**

Regression for prediction focuses on minimizing prediction error and may include many variables or interactions without concern for causal interpretation. In contrast, causal inference aims to estimate the effect of one variable on another, carefully considering confounding/omitted variables.

### 6.5 Algorithmic fairness

**Algorithms like COMPAS are used in real courtrooms to inform bail and sentencing decisions.
(a) Using your simple OLS model (16 vars), compute the mean predicted recidivism probability separately for Black defendants and white defendants in the test set. Is there a
difference?**

```{r}
test$pred_ols <- predict(model, newdata = test)

black_mean_pred <- test |>
  filter(black == 1) |>
  summarise(mean_pred = mean(pred_ols, na.rm = TRUE))

white_mean_pred <- test |>
  filter(black == 0,
         hispanic == 0,
         asian == 0,
         other_race == 0,
         native_american == 0) |>
  summarise(mean_pred = mean(pred_ols, na.rm = TRUE))

black_mean_pred
white_mean_pred

black_pred <- test |>
  filter(black == 1) |>
  summarise(mean_pred = mean(pred_ols, na.rm = TRUE)) |>
  pull(mean_pred)

white_pred <- test |>
  filter(black == 0,
         hispanic == 0,
         asian == 0,
         other_race == 0,
         native_american == 0) |>
  summarise(mean_pred = mean(pred_ols, na.rm = TRUE)) |>
  pull(mean_pred)

black_pred - white_pred

```

The mean predicted recidivism for black defendants is 51% and the mean predicted recidivism for white defendants is approximately 38%. This is a probability difference of 13.3 percentage points.

**(b) Now compute the actual recidivism rate for each group. Compare the predictions to the actual rates.**

```{r}
recid_table <- tibble(
  group = c("Black defendants", "White defendants"),
  mean_predicted = c(black_pred, white_pred),
  actual_rate = c(test |>
                    filter(black == 1) |>
                    summarize(mean(recidivism, na.rm = TRUE)) |>
                    pull(),
                  test |>
                    filter(black == 0,
                           hispanic == 0,
                           asian == 0,
                           other_race == 0,
                           native_american == 0) |>
                    summarize(mean(recidivism, na.rm = TRUE)) |>
                    pull()))

recid_table 
```

The actual recidivism rate is 0.521 for Black defendants and 0.402 for white defendants. The model’s predictions are fairly close to the actual rates for both groups, although it slightly underpredicts recidivism for each group.

**(c) In 2–3 sentences: What are the tradeoffs of using prediction algorithms in criminal justice Reference something specific from your analysis.**

While using prediction algorithms in criminal justice may improve consistency and accuracy, as seen in this analysis where predicted recidivism rates closely match actual outcomes for both groups, it can also predict inaccurately. For example, the model predicts substantially higher risk for Black defendants (0.512) than for white defendants (0.379), reflecting and potentially reinforcing existing disparities. This highlights a key tradeoff: while these tools can be helpful, they may also raise fairness concerns when used in high-stakes decisions.
